Covariate analysis of CMC and HBCC matched RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN normlaisation)

### 2017-08-30 code/Rmd/CMC_HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Changed the logic of age and PMI matching criteria

p = ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2) 
p
```

Age match CMC and HBCC (per Dx.Tissue category)
```{r preprocess.data1}
# Filter all ACC samples and Other samples
tmp = COVARIATES %>%
  dplyr::group_by(Dx.Tissue, Study) %>%
  dplyr::summarise(mn = quantile(Age_of_Death, 0.25), mx = quantile(Age_of_Death, 0.75)) %>%
  dplyr::group_by(Dx.Tissue) %>%
  dplyr::summarise(mn = max(mn), mx = min(mx))
COVARIATES = COVARIATES %>%
  dplyr::filter((Dx.Tissue == tmp$Dx.Tissue[1] & Age_of_Death >= tmp$mn[1] & Age_of_Death <= tmp$mx[1]) |
                  (Dx.Tissue == tmp$Dx.Tissue[2] & Age_of_Death >= tmp$mn[2] & Age_of_Death <= tmp$mx[2]))

@@

p = ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2)
p
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
COUNT[is.na(COUNT)] = 0
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dlply(.(Dx.Tissue), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
    processed.counts$filteredExprMatrix$genes
  }, COUNT)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p

@@

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation
Based on different model selection methodology (as seen here syn8468116, syn9772932, and syn9774491), following variables are used as covariates: Institution, Gender, RIN, and IntronicRate.

NOTE: Dx.Tissue.Study is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)

```{r custom.normalisation, results='asis'}
# Primary variable of interest
COVARIATES$Dx.Tissue.Study = paste(COVARIATES$Dx.Tissue, COVARIATES$Study, sep='_') %>%
  factor
postAdjustCovars = c('Dx.Tissue.Study', 'Institution', 'Gender', 'RIN', 'IntronicRate')

writeLines(paste('Using following covariates in the model:',
                 paste(postAdjustCovars, collapse=', ')))

@@

p = DE %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(Comparison~., scales = 'free_y')
p = p + ggtitle('Differential Expression between Schizophrenia and Control')
p
```

@@

# Store differential expression results
write.table(all.diff.exp, file = 'CMC_HBCC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_HBCC_DLPFC_DiffExpression.tsv', 
name = 'Differential Expression Results (Dx.Tissue)', 
parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |

>>>>

p1 = ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2) 
p1
```

Age match CMC and HBCC (per Dx.Tissue category)
```{r preprocess.data1}
# Filter all ACC samples and Other samples
max.age = COVARIATES %>%
  dplyr::group_by(Dx.Tissue, Study) %>%
  dplyr::filter(Study == 'CMC') %>%
  dplyr::summarise(mn = quantile(Age_of_Death, 0.5))

COVARIATES = COVARIATES %>%
  dplyr::filter((Dx.Tissue == max.age$Dx.Tissue[1] & Age_of_Death <= max.age$mn[1]) |
                  (Dx.Tissue == max.age$Dx.Tissue[2] & Age_of_Death <= max.age$mn[2]))

PMI match CMC and HBCC (per Dx.Tissue category)
```{r preprocess.data2}
max.pmi = COVARIATES %>%
  dplyr::group_by(Dx.Tissue) %>%
  dplyr::filter(Study == 'HBCC') %>%
  dplyr::summarise(mn = quantile(PMI, 0.5))

COVARIATES = COVARIATES %>%
  dplyr::filter((Dx.Tissue == max.age$Dx.Tissue[1] & PMI <= max.pmi$mn[1]) |
                  (Dx.Tissue == max.age$Dx.Tissue[2] & PMI <= max.pmi$mn[2]))

@@

p1 = ggpubr::ggarrange(plotlist = p, ncol = 4, nrow = 2)
p1
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
COUNT[is.na(COUNT)] = 0
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dlply(.(Dx.Tissue), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
    processed.counts$filteredExprMatrix$genes
  }, COUNT)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= label), size=4, hjust=0)
p

@@

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Iterative Normalisation
Dx.Tissue.Study is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)
```{r iterative.normalisation}
# Primary variable of interest
COVARIATES$Dx.Tissue.Study = paste(COVARIATES$Dx.Tissue, COVARIATES$Study, sep = '_') %>%
  factor()
postAdjustCovars = c('Dx.Tissue.Study');
avoid.covariates = c('Dx.Tissue', 'Study', 'Individual_ID')

# Assign residual covariates
residualCovars = setdiff(preAdjustedSigCovars$significantCovars, 
                         c(postAdjustCovars, avoid.covariates))
residualSigCovars = preAdjustedSigCovars
covariatesEffects = preAdjustedSigCovars$Effects.significantCovars[residualCovars]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 20){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', ')))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights for dispersion control
  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F)
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$E,
                            design = DM1$design,
                            weights = VOOM.GENE_EXPRESSION$weights)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                CQN.GENE_EXPRESSION$E)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates),
                         c(postAdjustCovars, avoid.covariates))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, avoid.covariates))
  covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```

### Normalisation
```{r custom.normalisation, results='asis'}
writeLines(paste('Using following covariates in the model:',
                 paste(postAdjustCovars, collapse=', ')))

@@

p = DE %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(Comparison~., scales = 'free_y')
p = p + ggtitle('Differential Expression between Schizophrenia and Control')
p

all.diff.exp = list(Diagnosis.Study = DE)
all.fit = list(Diagnosis.Study = ADJUSTED.FIT)
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=10, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE, aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE, aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE, aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

multiplot(plotlist = pl, cols = 3)
```

### Differential expression analysis (SCZ - Control gender specific)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp1, fig.height=15, fig.width=20}
COVARIATES$Dx.Tissue.Study.Gender = paste(COVARIATES$Dx.Tissue.Study,
                                          COVARIATES$Gender,
                                          sep = '_') %>%
  factor()

postAdjustCovars = c('Dx.Tissue.Study.Gender', setdiff(postAdjustCovars, c('Dx.Tissue.Study', 'Gender')))
writeLines(paste('Using following covariates in the model:', paste(postAdjustCovars, collapse=', ')))

# Fit contrast
contrast = makeContrasts(contrasts=
                           c("Dx.Tissue.Study.GenderSCZ_DLPFC_CMC_Female-Dx.Tissue.Study.GenderControl_DLPFC_CMC_Female",
                             "Dx.Tissue.Study.GenderSCZ_DLPFC_CMC_Male-Dx.Tissue.Study.GenderControl_DLPFC_CMC_Male",
                             "Dx.Tissue.Study.GenderSCZ_DLPFC_HBCC_Female-Dx.Tissue.Study.GenderControl_DLPFC_HBCC_Female",
                             "Dx.Tissue.Study.GenderSCZ_DLPFC_HBCC_Male-Dx.Tissue.Study.GenderControl_DLPFC_HBCC_Male"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR) 
names(DE) = gsub('Dx.Tissue.Study.Gender', '', colnames(contrast))

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Direction = sign(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(Ensemble2HGNC) %>%
  left_join(GENE.LEN)
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

p = DE %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(Comparison~., scales = 'free_y')
p = p + ggtitle('Differential Expression between Schizophrenia and Control')
p

all.diff.exp = c(all.diff.exp, list(Diagnosis.Study.Gender = DE))
all.fit = c(all.fit, list(Diagnosis.Study.Gender = ADJUSTED.FIT))
```

@@

# Store differential expression results
all.diff.exp %>%
  rbindlist(use.names = T, fill = T) %>%
  write.table(file = 'CMC_HBCC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_HBCC_DLPFC_DiffExpression.tsv', 
name = 'Differential Expression Results (Dx.Tissue)', 
parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |
